---
title: "RNA-seq QC and Unsupervised Analysis Report"
output: html_document
---

# Overview

This report demonstrates a modular RNA-seq-style analysis pipeline
implemented in R. Simulated gene expression data are generated,
missingness is introduced, normalization is applied, and quality
control metrics are computed prior to exploratory dimensionality
reduction and clustering.

The goal is to demonstrate foundational bioinformatics data
processing and unsupervised analysis workflows.

## Setup

```{r}
knitr::opts_chunk$set(echo = TRUE)

source("../scripts/simulate_data.R")
source("../scripts/preprocessing.R")
source("../scripts/qc_analysis.R")
source("../scripts/dim_reduction.R")
source("../scripts/run_pipeline.R")

```

---

# Running the Pipeline

```{r}
res <- run_pipeline(
  num_genes = 2000,
  num_samples = 30,
  missing_frac = 0.1,
  seed = 1,
  k = 3
)

```

---

# Overview of Simulated Data

Gene expression values were simulated from a log-normal distribution and 10% missingness was introduced at random.

```{r}
summary(res$expr_norm)

```

---

# QC Metrics

```{r}
head(res$sample_qc)
summary(res$sample_qc$missing_frac)

```

Missingness is centered around 10%, consistent with the simulation design.
Detection rates per sample are approximately 90%, suggesting uniform data
quality across samples. No samples exhibit extreme missingness or markedly
reduced detection rates, indicating that downstream dimensionality
reduction is unlikely to be driven by technical artifacts.

---

# PCA Results

```{r}
summary(res$pca)

plot_pca(res$pca)

```

The first two principal components explain X% and Y% of total variance,
respectively. These components represent the dominant axes of variation
across samples. Visual inspection of the PCA plot does not reveal extreme
outliers or clear separation patterns, which is expected given the
randomly simulated data.

---

# k-means Clustering

```{r}
res$kmeans$cluster

plot_kmeans_on_pca(res$pca, res$kmeans)

```

k-means clustering partitions samples into three groups based on similarity in principal component space. Given the simulated nature of the data without embedded biological structure, clusters reflect random structure rather than meaningful biological grouping.

---

# Potential Next Steps

- Model biologically realistic dropout by increasing missingness probability
  for low-expression genes.
- Implement additional outlier detection metrics, such as sample correlation
  thresholds or variance-based filtering.
- Incorporate gene filtering prior to PCA to remove low-detection genes.
- Compare clustering performed directly on normalized data versus PCA-reduced space.
